---
title: rutils Package for Simplifying Financial Data Management and Modeling
author: Jerzy Pawlowski (algoquant)
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"rutils Package"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The *rutils* package is a library of functions designed for simplifying data management and data modeling, and is used by many other algoquant packages.  
The *rutils* package contains functions for:

+ extracting names and columns from time series, 
+ calculating the end points of a time series, 
+ applying lags to vectors and matrices, 
+ calculating row differences of vectors and matrices, 
+ recursively binding a list of objects into time series, 
+ recursively applying a function to a list of objects, 
+ plotting time series with custom axis range and background shading. 

The *rutils* package also includes a dataset with *OHLC* time series data for a portfolio of symbols.  The data is contained in an environment called *env_etf*, which includes: 

+ *sym_bols*: a vector of strings with the portfolio symbols.
+ individual time series *VTI*, *VEU*, etc., containing daily *OHLC* prices for all the sym_bols.
+ *price_s*: a single *xts* time series containing daily closing prices for all the sym_bols.
+ *re_turns*: a single *xts* time series containing daily returns for all the sym_bols.

Each individual *xts* time series contains the columns: Open prices, High prices, Low prices, Close prices, trading Volume, Adjusted prices.


========

### Installation and loading

Installing *rutils* package from *GitHub*:  
```{r eval=FALSE}
install.packages("devtools")
library(devtools)
install_github(repo="algoquant/rutils")
library(rutils)
```

========

### Data

The *rutils* package contains a dataset of daily *OHLC* time series in *xts* format, for a portfolio of stock symbols.  The time series are contained in an environment called *env_etf*.  The data is set up for lazy loading, so it doesn't require calling `data(etf_data)` to load it before being able to call it.  
```{r echo=-1, eval=FALSE}
library(rutils)
# get first six rows of VTI prices
head(env_etf$VTI)
# plot
chart_Series(x=env_etf$VTI["2009-11"])
```


========

### Examples

Extract the name of an *OHLC* time series from its first column name:  
```{r echo=-1, eval=TRUE}
suppressMessages(suppressWarnings(library(rutils)))
# show portfolio symbols
env_etf$sym_bols
# get name for VTI
na_me(env_etf$VTI)
# get first six rows of VTI prices
head(env_etf$VTI)
# get first six rows of price_s
env_etf$price_s[1:6, 1:4]
# get first six rows of re_turns
env_etf$re_turns[1:6, 1:4]
```


Calculate a vector of equally spaced end points for a time
series:  
```{r eval=FALSE}
# calculate end points with initial stub interval
end_points(env_etf$VTI, inter_val=7, off_set=4)
```


Extract close prices from an *OHLC* time series:  
```{r eval=FALSE}
# get close prices for VTI
clo_se(env_etf$VTI)
# get volumes for VTI
clo_se(env_etf$VTI, which_col="vol")
```


Apply a lag to a vector or matrix:  
```{r eval=FALSE}
# lag vector by 2 periods
lag_it(1:10, lag=2)
# lag matrix by negative 2 periods
lag_it(matrix(1:10, ncol=2), lag=-2)
```


Calculate the row differences of a vector or matrix:  
```{r eval=FALSE}
# diff vector by 2 periods
diff_it(1:10, lag=2)
# diff matrix by negative 2 periods
diff_it(matrix(1:10, ncol=2), lag=-2)
```


Calculate the time differences of an *xts* time series and pad with zeros:  
```{r eval=FALSE}
# calculate time differences over lag by 10 periods
rutils::diff_xts(env_etf$VTI, lag=10)
```


Recursively rbind a list of objects:  
```{r eval=FALSE}
# create xts time series
x_ts <- xts(x=rnorm(1000), order.by=(Sys.time()-3600*(1:1000)))
# split time series into daily list
list_xts <- split(x_ts, "days")
# rbind the list back into a time series and compare with the original
identical(x_ts, do_call_rbind(list_xts))
```


Recursively apply a function to a list of objects:  
```{r eval=FALSE}
# create xts time series
x_ts <- xts(x=rnorm(1000), order.by=(Sys.time()-3600*(1:1000)))
# split time series into daily list
list_xts <- split(x_ts, "days")
# rbind the list back into a time series and compare with the original
identical(x_ts, do_call(rbind, list_xts))
```


Apply a function to a list of objects, merge the outputs into a single object, and assign the object to the output environment:  
```{r eval=FALSE}
do_call_assign(
   func_tion=clo_se,
   sym_bols=env_etf$sym_bols,
   out_put="price_s",
   env_in=env_etf, env_out=new_env)
```


Plot an *xts* time series with custom y-axis range and with vertical background shading:  
```{r echo=-1, eval=TRUE, fig.width=6, fig.height=4}
suppressMessages(suppressWarnings(library(rutils)))
chart_xts(env_etf$VTI["2015-11"],
name="VTI in Nov 2015", ylim=c(102, 108),
in_dex=index(env_etf$VTI["2015-11"]) > as.Date("2015-11-18"))
```

